---
title: "Practicum I CS5200"
author: "Krina Devani (devani.k@northeastern.edu), Sarthak Kagliwal (kagliwal.s@northeastern.edu)"
date: "Spring 2023"
output:
  pdf_document: default
  html_document:
    df_print: paged
---



##Connect to Database
This Practicum will use a local instance of MySQL. This database is very easy to set up and with minimal configuration or management.
To connect the database, we are using 'dbConnect()' function which takes in the username, password and database configurations.

```{r dbConnectBlock, eval = T, warning = F}

# 1. Library 
library(RMySQL)
library(DBI)
library(sqldf)
library(plyr)
library(dplyr)
library(ggplot2)
library(tinytex)

# 2. Settings
db_user <- 'root'
db_password <- 'Bkjbkz@403'
db_name <- 'dbtest'
db_host <- '127.0.0.1' 
db_port <- 3306 

# 3. Connect to DB
mydb <-  dbConnect(MySQL(), user = db_user, password = db_password,
                 dbname = db_name, host = db_host, port = db_port)

```


To check if the database connection is successful. 
Returns TRUE if connected successfully, otherwise FALSE.
```{r testDbConnection, eval = T, warning = F}
dbIsValid(mydb)
```


##Create Database

#4A. Create table incidents
This table will stores wildlife strike incidents.

```{sql dropIncidentsTable, connection=mydb}

DROP TABLE IF EXISTS incidents;

```

Writing the 'incidents' table schema.
Here rid is the primary key. 
'origin' is a Foreign key(FK) which refrences the airport table.
'airline' is a Foreign key(FK) which refrences the airline table.
'conditions' is a Foreign key(FK) which refrences the conditions table.
'flight.phase' is a value from the value set ("takeoff", "landing", "inflight", "unknown")
```{sql createIncidentsTable, connection=mydb}

CREATE TABLE incidents(
  rid INTEGER NOT NULL  PRIMARY KEY,
  `dep.date` DATE NULL,
  origin INTEGER NULL,
  airline INTEGER NULL,
  aircraft TEXT NULL,
  `flight.phase` SET ("takeoff", "landing", "inflight", "unknown") NOT NULL DEFAULT 'unknown',
  altitude INTEGER NULL,
  conditions VARCHAR(255) NULL,
  warned BOOLEAN NOT NULL DEFAULT FALSE,
  CHECK(altitude >= 0)
)

```


#4B. Create table airports
This table will stores airport name, airport code and states.

```{sql dropAirportsTable, connection=mydb}

DROP TABLE IF EXISTS airports;

```


Writing the 'airports' table schema.
Here aid is the primary key. 
```{sql createAirportsTable, connection=mydb}

CREATE TABLE airports(
  aid INTEGER NOT NULL PRIMARY KEY AUTO_INCREMENT,
  airportName TEXT NOT NULL,
  airportCode TEXT NULL,
  state TEXT
)

```


#4C. Setup origin foreign key
Given chunk will link the incidents and airports tables via the origin foreign key in incidents to the primary key aid in airports.

```{sql addFkOriginBlock, connection=mydb}

ALTER TABLE incidents ADD CONSTRAINT airports_origin_fkey FOREIGN KEY (origin) REFERENCES airports(aid);

```


#4D. Create lookup table conditions
This table will link incidents table with conditions foreign key.

```{sql dropConditionsTable, connection=mydb}

DROP TABLE IF EXISTS conditions;

```

Writing the 'conditions' table schema.
Here cid is the primary key. 
Each condition value in a row is unique.
```{sql createConditionsTable, connection=mydb}

CREATE TABLE conditions(
  cid INTEGER NOT NULL PRIMARY KEY AUTO_INCREMENT,
  `condition` VARCHAR(255) NOT NULL,
  explanation TEXT NULL,
  UNIQUE (`condition`)
)

```

Make conditions column a foreign key.
```{sql alterConditionsToInt, connection=mydb}

ALTER TABLE incidents MODIFY conditions INTEGER;

```


Given chunk will link the incidents and conditions tables via the conditions foreign key in incidents to the primary key cid in condtions.
```{sql addFkConditionsBlock, connection=mydb}

ALTER TABLE incidents ADD CONSTRAINT conditions_fkey FOREIGN KEY (conditions) REFERENCES conditions(cid);

```


#4E. Create table airlines
This table will stores airlines name, airlines code and flag.

```{sql dropAirlinesTable, connection=mydb}
DROP TABLE IF EXISTS airlines;
```


Writing the 'airlines' table schema.
Here eid is the primary key. 
```{sql createAirlineTable, connection=mydb}
CREATE TABLE airlines(
  eid INTEGER NOT NULL PRIMARY KEY AUTO_INCREMENT,
  airlineName TEXT NOT NULL,
  airlineCode TEXT NULL,
  flag TEXT NULL
)
```


#4F. Setup origin foreign key
Given chunk will link the incidents and airlines tables via the airline foreign key in incidents to the primary key eid in airlines.

```{sql addFkAirlineBlock, connection=mydb}

ALTER TABLE incidents ADD CONSTRAINT airports_airlines_fkey FOREIGN KEY (airline) REFERENCES airlines(eid);

```


This chunk will be used to test the table definitions.
```{r evaluateTableDef, eval = F, warning = F}

dbGetQuery(mydb, "DESC incidents")
dbGetQuery(mydb, "DESC airports")
dbGetQuery(mydb, "DESC conditions")
dbGetQuery(mydb, "DESC airlines")

```


##Load CSV data

#Load CSV file into the dataframe
Given chunk will load data from given BirdStrikesData-V2.csv to the bds.raw dataframe.

```{r loadCSVToDataframe, eval = T, warning = F}
bds.raw <- read.csv("BirdStrikesData-V2.csv", stringsAsFactors = FALSE)
```


To Check if the csv file is loaded in the dataframe.
```{r populateDataframe, eval = T, warning = F}
head(bds.raw)
```

Map the flight phase given in the csv to the value set.
```{r mapFlightPhase, eval = T, warning = F}

x <- c("Parked","Taxi","Take-off run","Climb", "Approach", "Descent", "Landing Roll","")
y <- c("unknown","unknown","takeoff","takeoff","inflight","landing","landing","unknown")
bds.raw$flight_phase <- mapvalues(bds.raw$flight_phase, 
          from=x, 
          to=y)

```

Set default value to flight date.
```{r setDefalutToFlightDate, eval = T, warning = F}

bds.raw$flight_date[bds.raw$flight_date == "" | bds.raw$flight_date == "N/A" | bds.raw$flight_date == "UNKNOWN" | is.na(bds.raw$flight_date) == TRUE ] <- "1/1/2000 0:00"

```


Set default value to airport & origin and get their distinct value and store the values in a subset of bds.raw dataframe 'view1'.
```{r setDefaultToAirportOrigin, eval = T, warning = F}

view1 <- distinct(bds.raw[,c(3,10)],.keep_all=TRUE)
view1$airport[view1$airport == "" | view1$airport == "N/A" | view1$airport == "UNKNOWN" | is.na(view1$airport) == "TRUE"] <- "unknown" 
view1$origin[view1$origin == "" | view1$origin == "N/A" | view1$origin == "UNKNOWN" | is.na(view1$origin) == "TRUE"] <- "unknown"
view1

```


#Load data to airports table
Given chunk will load data from data frame to airports table.

```{r loadAiportTable, eval = T, warning = F}

query <- paste0("INSERT INTO airports (airportName, state) VALUES ", 
                paste(sprintf('("%s", "%s")', view1$airport, view1$origin), 
                      collapse = ","))

dbGetQuery(mydb, query)

```


Updating the view1 dataframe with the data stored in the airport table with its primary key.
```{r updateView1, eval = T, warning = F}

view1 <- dbGetQuery(mydb,"SELECT aid, airportName AS airport, state as stateVal FROM airports")
```

Testing the view1 dataframe.
```{r testView1, eval = T, warning=F}
head(view1)
```


Set default value to airline and get its distinct value and store the values in a subset of bds.raw dataframe 'view2'.
```{r setDefaultToAirline, eval = T, warning = F}

view2 <- distinct(bds.raw['airline'],.keep_all=TRUE)
view2$airline[view2$airline == "" | view2$airline == "N/A" | view2$airline == "UNKNOWN" | is.na(view2$airline) == "TRUE"] <- "unknown" 
view2

```


#Load data to airlines table
Given chunk will load data from data frame to airlines table.

```{r loadAirlinesTable, eval = T, warning = F}

query <- paste0("INSERT INTO airlines (airlineName) VALUES ", 
                paste(sprintf('("%s")', view2$airline), 
                      collapse = ","))

dbExecute(mydb, query)

```
Updating the view2 dataframe with the data stored in the airlines table with its primary key.
```{r updateView2, eval = T, warning = F}

view2 <- dbGetQuery(mydb,"SELECT eid, airlineName AS airline FROM airlines")

```

Testing the view2 dataframe.
```{r testView2, eval = T, warning=F}

head(view2)

```


Get distinct value of sky conditions.
```{r}

view3 <- distinct(bds.raw["sky_conditions"],.keep_all = TRUE)
view3

```


#Load data to conditions table
Given chunk will load data from data frame to conditions table.

```{r loadConditionsTable, eval = T, warning = F}

query <- paste0("INSERT INTO conditions (`condition`) VALUES ", 
                paste(sprintf("('%s')", view3$sky_conditions), 
                      collapse = ","))

dbExecute(mydb, query)

```
Updating the view3 dataframe with the data stored in the conditions table with its primary key.
```{r  updateView3, eval = T, warning = F}

view3 <- dbGetQuery(mydb,"SELECT cid,`condition` AS sky_conditions FROM conditions")

```

Testing the view3 dataframe.
```{r testView3, eval = T, warning=F}
head(view3)
```

Make default unknown when airport and origin is empty.
```{r createDemoDataframe, eval = T, warning = F}

demo<- bds.raw
demo$airport[demo$airport == "" | demo$airport == "N/A" | demo$airport == "UNKNOWN" | is.na(demo$airport) == "TRUE"] <- "unknown" 
demo$origin[demo$origin == "" | demo$origin == "N/A" | demo$origin == "UNKNOWN" | is.na(demo$origin) == "TRUE"] <- "unknown"
unique(demo$airport[demo$airport == "unknown"])

```

Apply left joins on view1, view2 and view3 and create new dataframe temp.
```{r makeDfUsingJoins, eval = T, warning = F}

temp <-left_join(left_join(left_join(demo,view1,by = c('airport', 'origin'='stateVal')),view2, by = "airline"),view3, by = "sky_conditions")
temp
```

Create a new dataframe 'view4' that contains data and columns related to incidents table.
```{r formatFlightDate, eval = T, warning = F}

view4 <- temp[,c(1,7,20,21,2,11,18,22,17)]

view4$flight_date <- gsub('-', '/', view4$flight_date)
view4$flight_date <- as.Date(view4$flight_date, "%m/%d/%y")

view4
```


#Load data to incidents table
Given chunk will load data from data frame to incidents table.

```{r loadIncidentsTable, eval = T, warning = F}

query <- paste0("INSERT INTO incidents (rid, `dep.date`, origin, airline, aircraft, `flight.phase`, altitude, `conditions`, warned) VALUES ", 
                paste(sprintf("('%s', NULLIF('%s', 'NA'), NULLIF('%d', 'NA'), NULLIF('%d', 'NA'),'%s', '%s', NULLIF('%s', 'NA'), NULLIF('%d', 'NA'), '%d')",
                              view4$rid, view4$flight_date, view4$aid, view4$eid, view4$aircraft, view4$flight_phase, as.numeric(gsub(",", "", view4$altitude_ft)), view4$cid, ifelse(view4$pilot_warned_flag == 'Y', TRUE, FALSE)),
                      collapse = ","))


dbExecute(mydb, query)

```


#Display data loaded in the table

Display incidents table data
```{r displayIncidentsData, eval = T, warning = F}

query <- "SELECT * FROM incidents LIMIT 10"
dbGetQuery(mydb, query)

```


Display airports table data
```{r displayAirportsData, eval = T, warning = F}

query <- "SELECT * FROM airports LIMIT 10"
dbGetQuery(mydb, query)

```

Display airlines table data
```{r displayAirlinesData, eval = T, warning = F}

query <- "SELECT * FROM airlines LIMIT 10"
dbGetQuery(mydb, query)

```

Display conditions table data
```{r displayConditionsData, eval = T, warning = F}

query <- "SELECT * FROM conditions LIMIT 10"
dbGetQuery(mydb, query)

```


##Queries

#8 
SQL query to find 10 states with the greatest number of incidents.

```{sql findStateWithIncidents, connection=mydb}

SELECT 
  ar.state, 
  COUNT(*) NoOfIncidents 
FROM incidents AS i 
LEFT JOIN airports AS ar 
  ON i.origin = ar.aid 
  GROUP BY ar.state 
ORDER BY NoOfIncidents DESC 
LIMIT 10

```


#9
SQL query against your database to find the airlines that had an above average number bird strike incidents

```{sql getAirlineAboveAvgIncidents, connection=mydb}

SELECT 
  ai.airlineName, 
  COUNT(*) NoOfIncidents
FROM incidents AS i
LEFT JOIN airlines AS ai 
  ON i.airline = ai.eid
  GROUP BY ai.airlineName
HAVING NoOfIncidents > (
    SELECT 
      AVG(NoOfIncidents)
    FROM (
    SELECT 
      ai.airlineName, 
      COUNT(*) NoOfIncidents
    FROM incidents AS i
    LEFT JOIN airlines AS ai
        ON i.airline = ai.eid
        GROUP BY ai.airlineName
    )AS av
)
ORDER BY NoOfIncidents DESC

```


#10
Create a SQL query against your database to find the number of bird strike incidents by month and by flight phase (across all years).

```{r getIncidentsByMonthAndPhase, eval = T, warning = F}

query <-"SELECT MONTH(i.`dep.date`) AS Month, i.`flight.phase` AS flightPhase, COUNT(*) NoOfIncidents FROM incidents AS i GROUP BY MONTH(i.`dep.date`), i.`flight.phase` ORDER BY Month, NoOfIncidents;"

dVal <- dbGetQuery(mydb,query)
head(dVal)

```


#11
Scatter plot that plots month along the x-axis versus number of incidents.

```{r scatterPlot, eval = T, warning = F}

sumDf <- aggregate(dVal['NoOfIncidents'], list(Month = dVal$Month),sum)
plot(sumDf$Month,
     sumDf$NoOfIncidents, 
     xlab = "Months", 
     ylab = "No of incidents", 
     main = "Bird strike incidents by month",
     pch = 16,
     label = sumDf$NoOfIncidents,
)
legend("topleft",
      legend = c("NoOfIncidents"),
      col = c(rgb(0,0,0)),
      pch = c(16),
      inset = c(0.1)
      )
text(sumDf$Month,
     sumDf$NoOfIncidents+100, labels=sumDf$NoOfIncidents)

```


#12
Given stored procedure will add a new incident to the database.

```{sql dropSPAddNewIncident, connection=mydb}
DROP PROCEDURE SPAddNewIncident;
```

```{sql sPAddNewIncident, connection=mydb}

CREATE PROCEDURE SPAddNewIncident(
  nRid INTEGER,
  nDepDate DATE,
  nAirport TEXT,
  nOrigin TEXT,
  nAirline TEXT,
  nAircraft TEXT,
  nPhase TEXT,
  nConditions TEXT,
  nAltitude INTEGER,
  nWarned BOOLEAN
)
BEGIN

DECLARE naid INT DEFAULT 0;
DECLARE neid INT DEFAULT 0;
DECLARE ncid INT DEFAULT 0;

IF (ISNULL(nRid))
THEN
SIGNAL SQLSTATE '42601'
SET MESSAGE_TEXT = 'rid can not be null!';
END IF;

IF (nAltitude < 0)
THEN
SIGNAL SQLSTATE '42601'
SET MESSAGE_TEXT = 'altitude can not be negative!';
END IF;

IF (nPhase not in ("takeoff", "landing", "inflight", "unknown"))
THEN
SIGNAL SQLSTATE '42601'
SET MESSAGE_TEXT = 'phase is incorrect!';
END IF;


IF (nAltitude >= 0)
THEN
  
    SET naid = (SELECT aid FROM airports WHERE airportName = nAirport AND state = nOrigin);
    IF (ISNULL(naid))
    THEN 
    
      INSERT INTO airports (airportName, state) VALUES (nAirport, nOrigin);
      SET naid = (SELECT aid FROM airports WHERE airportName = nAirport AND state = nOrigin);
      
    END IF;
    
    
    SET neid = (SELECT eid FROM airlines WHERE airlineName = nAirline);
    IF (ISNULL(neid))
    THEN 
    
      INSERT INTO airlines (airlineName) VALUES (nAirline);
      SET neid = (SELECT eid FROM airlines WHERE airlineName = nAirline);
      
    END IF;
  
    
    SET ncid = (SELECT cid FROM conditions WHERE `condition` = nConditions);
    
    INSERT INTO incidents (rid, `dep.date`, origin, airline, aircraft, `flight.phase`, altitude, `conditions`, warned)
    VALUES
    (nRid, nDepDate, naid, neid, nAircraft, nPhase, nAltitude, ncid, nWarned);
    
END IF;
  
END

```


Using the stored procedure 'SPAddNewIncident' to add a new incident.
```{sql testSPAddNewIncident,connection=mydb}
CALL SPAddNewIncident(11, '1997-06-10','LAGUARDIA NY', 'NY', 'AMERICAN AIRLINES', 'Airplane', 'unknown', 'No Cloud', 50, TRUE);
```

Checking if the Stored procedure worked correctly.
```{sql checkSP ,connection=mydb}

select * from incidents where rid = 11

```

Test stored procedure if provided rid is null.
It will raise concern while adding new data.
```{sql testSPRidNull,connection=mydb}
CALL SPAddNewIncident(null, '1997-06-10','LAGUARDIA NY', 'NY', 'AMERICAN AIRLINES', 'Airplane', 'unknown', 'No Cloud', 50, TRUE);
```

##Disconnect to Database
```{r dbDisconnectBlock, eval = T, warning = F}

# Disconnect database connection
dbDisconnect(mydb)

```




